---
title: "Present a great story for data science projects"
subtitle: "This is a Slide Template"
author: "Namita Mishra (Advisor: Dr. Ashraf Cohen)"
date: '`r Sys.Date()`'
format:
  revealjs
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!

More information about `revealjs`:
<https://quarto.org/docs/reference/formats/presentations/revealjs.html>
:::

## Introduction {.smaller}

Type 2 Diabetes (T2D), a major public health concern worldwide prompts
us to understand the factors contributing to diabetes risk guiding us to
develop effective prevention strategies. With diabetes, the body doesn’t
make enough insulin or can’t use the insulin it makes and causes too
much blood sugar to stay in the bloodstream posing risks of having heart
disease and other organ damage. Recognizing risk factors of diabetes
could help prevent DM as an early intervention. Research finds - Age,
body mass index (BMI), and race as risk factors. Modeling their
interactions and effects presents methodological challenges. Traditional
statistical approaches fail to analyze complex relationships or
uncertainty in predictions. Here, we address the challenges of
healthcare data analysis using both Frequentist and Bayesian methods on
the NHANES dataset\> the study is an attempt to identify association of
these risk factors with diabetes and measure the predictive performance
of the model, and demonstrate methodological perspectives that could
enhance actionable insights in healthcare.

## Methods

Dataset (NHANES)

| Dataset | Key variables    | Col3       |
|---------|------------------|------------|
| demo1   | age,race, gender | 10175 × 47 |
| exam1   | BMI              | 9813 × 224 |
| quest1  | Diabetes         | 9813 × 953 |

-   In this model, we analyze the probability of diabetes based on BMI,
    age, and race in the dataset with a small sample for certain
    subgroups in the BMI category, where frequentist estimates are
    unstable.

-   Bayesian regression allows the incorporation of prior knowledge and
    improves estimates.

-   Posterior distributions provide a full uncertainty measure as
    credible intervals rather than just point estimates.

-   Data with missing values are handled in Bayesian methods by aturally
    incorporate uncertainty.

-   The predictors show relationships and interactions that are complex
    or hierarchical in structure. In the dataset the the pre-diabetic
    category is underrepresented (n=132/9813)

-   a Bayesian regression seem appropriate that would allow
    incorporation of prior knowledge to stabilize estimates, can account
    for uncertainty due to missing values, and provides full posterior
    distributions, offering robust inference than traditional
    frequentist methods for this dataset.

## Data Exploration and Visualization

-   NHANES (CDC), .xpt files imported using r

-   Findings:

    -   High proportion of missing data (BMI and diabetes status) and
        has imbalance in observed diabetes outcomes (Yes \> No). Age
        range of (0,- 80) with median and mean age (29 and 32.5 resp.)
        shows very young participants (0–10 years) in the
        diabetes-focused study is unusual, considering the prior
        knowledge that diabetes is rare in children. Survey-weighted
        proportions reflects population estimates Male: female :: 48.9%:
        51%, with the majority participants fall into Non-Hispanic White
        category. Most adults are of normal weight, and the weighted
        proportions are suitable to run Bayesian regression that
        represents the US population.

## Data Exploration and Visualization {.smaller}

A study was conducted to determine how...

```{r, warning=FALSE, echo=F, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=F}
# Load Data
#kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

## Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

## Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
